---
title: "SMLP_Iliopoulos"
author: "Ioannis Iliopoulos"
date: "2023-09-09"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Revision

RK comments added.

## EEG dataset

The following dataset includes pre-processed EEG data from a total of 105 participants who took part in a semantic violation experiment. 
The dependent variable is to be found under the variable named "value". The values there are already aggregated for the time window 350-600 ms after stimulus onset.
The rest of the variables include:
"subject": Participant Number
"item_n": Item Number 
"Condition": Two Conditions: 1 (semantically plausible) and 9 (semantically implausible)
"Animacy": This was an extra manipulation that I still haven't looked at. We checked verb-noun combinations and this variable refers to the animacy status of the noun. 
"variable": This variable includes the name of the EEG-channels. Depending on the analysis, I always keep only a subset of it. 

```{r cars}
data<-read.csv("data/sem_de_350_600.csv")
summary(data)
```

## Analysis

For this analysis I just keep the Channels where the N400 is expected ("Cz","CPz","CP1","CP2","Pz"). I use the condition 1 (semantically plausible) as baseline condition, not using any contrast coding. 

```{r pressure}
library(lme4)
library(tidyverse)

individual_analysis_channels<-c("Cz","CPz","CP1","CP2","Pz")
sem_de_individual <- droplevels(subset(data, variable %in% individual_analysis_channels))
sem_N400_data<-aggregate(value~condition+subject+item_n,data=data, FUN = mean)
sem_N400_data$condition<-as.factor(sem_N400_data$condition)

m1 <- lmer(value ~ condition + (1+condition|subject)+(1|item_n), data = sem_N400_data)
isSingular(m1)     # ok
summary(rePCA(m1)) # ok
VarCorr(m1)

# RK -- we usuallly use REML=FALSE; also control=lmerControl(calc.derivs=FALSE)
m1a <- lmer(value ~ condition + (1+condition|subject)+(1|item_n), data = sem_N400_data, REML=FALSE)
isSingular(m1a)     # not ok
summary(rePCA(m1a)) # not ok
VarCorr(m1a)
```

RK comment:

1. There is a problems with your data representation: `subject` and `item_n` need to be factors. 

2. `condition` is both within-subject and within-item. So you may also want check for corresponding item-related VCs and CPs. I set up the data in my default style. This actually leads to a case where `REML=TRUE` (lme4 default) misses a degenerate fit. 
3. Recommended default lme4 specification uses `REML=FALSE` and `control=lmerControl(calc.derivs=FALSE)`.

I will work with my default specification and variable-naming style to illustrate solutions for some of what you write in _Open Issues_.

```{r revision}
dat <- 
  sem_N400_data |> 
  as_tibble() |> 
  mutate(Subj = as_factor(paste0("S", str_pad(subject, width = 3, side = "left", pad = "0"))),
         Item = as_factor(paste0("I", str_pad(item_n, width = 2, side = "left", pad = "0"))),
         Cond = fct_recode(condition, "plausible" = '1', "implausible" = '9')) |> 
  select(Subj, Item, Cond, value) |> 
  arrange(Subj, Item, Cond)

# check distribution? 

# 0 Set contrasts explicitly
contrasts(dat$Cond) <- contr.treatment(2)

# 1  Recommended default specification
m2 <-lmer(value ~ Cond + (1+Cond | Subj) + (1 + Cond | Item), data = dat, REML=FALSE,
          control=lmerControl(calc.derivs=FALSE))
isSingular(m2)     # not ok
summary(rePCA(m2)) # not ok
VarCorr(m2)

# 2  Extract indicator variable from model matrix 
mm <- model.matrix(~ 1 + Cond, data=dat)
dat$cond <- mm[, 2]

# 3 Fit zero-correlation parameter LMM
m3 <-lmer(value ~ Cond + (1+cond || Subj) + (1 + cond || Item), data = dat, REML=FALSE,
          control=lmerControl(calc.derivs=FALSE))
isSingular(m3)     # not ok
summary(rePCA(m3)) # not ok
VarCorr(m3)

# 4 Fit only varying intercepts LMM for control
m4 <-lmer(value ~ Cond + (1 | Subj) + (1 | Item), data = dat, REML=FALSE,
          control=lmerControl(calc.derivs=FALSE))

anova(m4, m3, m2)
```

RK

1. Your baseline is collinear with the condition effect, both for subjects and items. 
2. There is no evidence for reliable between-subjects differences in the condition effect.
 
## Open Issues

1. I am interested in individual variability and more specifically within-person variability. For that I have tested the same individuals in 2 different EEG experiments in their first and their second language. Concretely, I am aiming to find differences in the L1 and the L2 concerning N400 and P600. For that I usually extract the slopes by participant and compare them (using correlation tests) with the slopes from the other experiments. What can I do when adding the random slopes gives me a boundary warning? Is there a better way to check for this?

RK: I would need to see the data. There are a few options and we will definitely cover this at SMLP2023. Mainly:

+ Make sure the boundary warning is not a false positive. Use the `control=lmerControl(calc.derivs=FALSE)` argument (see above and below).
+ Aim for a parsimonious LMM, that is remove variance components (VCs) and correlation parameter (CPs) not supported by the data.

2. (1+condition|subject) VS (1+condition||subject). Why when extracting the random slopes I get an extra column when the intercept and slope by participant are uncorrelated? How do I solve this? \

+ You need to switch from factors to indicator variables. See above In MixedModels.jl you use zerocorr() for factors. 

We save the data for input to Julia. 

```{r output}
library(arrow)
write_feather(dat, "data/Iliopoulos_N400.arrow")
```


# Appendix

```{r}
sessionInfo()
```

